### 2012-10-08 Doc/VRIPHYS2012/vriphys2012.tex
final version of VRIPHYS paper
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@463 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

% --- for  Annual CONFERENCE
% \ConferenceSubmission % uncomment for Conference submission
% \ConferencePaper      % uncomment for (final) Conference Paper
% \STAR                 % uncomment for STAR contribution
% \Tutorial             % uncomment for Tutorial contribution
% \ShortPresentation    % uncomment for (final) Short Conference Presentation
%
% --- for  CGF Journal
% \JournalSubmission    % uncomment for submission to Computer Graphics Forum
% \JournalPaper         % uncomment for final version of Journal Paper
%
% --- for  EG Workshop Proceedings
\WsSubmission    % uncomment for submission to EG Workshop
% \WsPaper         % uncomment for final version of EG Workshop contribution
%
 \electronicVersion % can be used both for the printed and electronic version

@@

@@

% for anonymous conference submission please enter your SUBMISSION ID
% instead of the author's name (and leave the affiliation blank) !!
% TODO: anonymize -- Submission ID
       \author[paper1026]
              {paper1026}
%\author[T. Golembiovský \& C. Duriez]
%       {Tomáš Golembiovský$^{1,2}$ and Christian Duriez$^{2}$\\
%        $^1$Faculty of Informatics, Masaryk University, Czech Republic\\
%        $^2$INRIA Lille, France
%       } % TODO: ask Christian how to specify affiliation

@@

\begin{classification} % according to http://www.acm.org/class/1998/
    \CCScat{Computer Graphics}{I.3.5}{Computational Geometry and Object Modeling}{Physically based modelling}.
    %\CCScat{Computer Graphics}{I.3.7}{Three-Dimensional Graphics and Realism}{Animation}.
    %\CCScat{Simulation and Modeling}{I.6.8}{Types of Simulation}{Animation}
    %\CCScat{Life and Medical Sciences}{J.3}{Health}{}
\end{classification}

@@

Thin objects are very frequent in our every-day life. % thin wall objects ?
%Objects with thin structure occur in many forms 
e.g. textiles, paper, leaves etc. 
Such objects are also very common in anatomical or pathological structures for example tubular structures (like blood vessels, colon \ldots), thin bag structures (%vessie, 
Glisson capsule, aneurysms,  \ldots), and also many others (eyes, skin, \ldots). 
In the field of surgical simulation, we try to reproduce or anticipate the mechanical behavior of these structures during the surgery.
Thus, it is very important to capture accurately the physical behavior of these structure.
However, in the same time, it is also very important to maintain fast computation rates, in order to be compatible with the targeted applications (interactive simulation for education and for planning).

This problem has been faced by the computer graphics community but for different applications.
For instance, many related works address the problem of the simulation of cloth using mass-spring models \cite{Volino2009} or bending models \cite{Grinspun2003}. %% Todo Choi2007?
However, these models are discrete so their behavior depends on the mesh and their parameters are not easily measurable.
This is completely acceptable for some computer graphics application, where the parameters can be tuned manually to obtain a realistic animation.
In the case of medical simulation, more physically based modeling is needed.

Methods based on continuum mechanics and namely theory of elasticity gained on popularity in real-time simulations ever since Terzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.
While the methods using finite elements and theory of elasticity are fairly popular in volumetric modeling of deformations \cite{Cotin1999,Nesme2006,Miller2007}, they are still frowned upon in the area of thin structures. 
Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
This theory was integrated numerically using Finite Elements \cite{Reddy1993}, but there are two issues for obtaining interactive frame rates: 
(i) the continuum equations requires second spatial derivations that can not be handled with linear interpolation and 
(ii) the final system of equations that describes the mechanics is non-linear (if the deformations are not small).

The usual argument against non-linear FEM based methods is that they are computationally too expensive. 
This is indeed true if lots of elements are used to discretize the object.
A classical strategy, is to use high-polygonal meshes for rendering or collision detection and response, while maintaining a reasonably low number of elements for the mechanics.
The work of Bouthors et al.\cite{Bouthors2007} is a good application example of this strategy for surface deformations.
But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements that have an interpolation with high order polynomials.
This type of interpolation is necessary because bending continuum model
(based on Kirchhoff-Love theory for thin plates)  requires the computation of second spatial derivatives.
On the other hand, elements common in mechanics (e.g. Discrete Kirchhoff Triangles) are based on shape functions that are impractical for real-time simulations or don't have explicitly defined shape functions at all.
In \cite{Comas2010c}, Comas et al. use a hybrid interpolation: linear interpolation is used for in-plane deformations and cubic polynomials for bending deformations.
The element is piloted by 3 nodes on which 6 degrees of freedom are defined. 
From this point of view, our work is derived from this approach but we address several noted drawbacks: the formulation misses one rotational degree of freedom, the cubic interpolation is not symmetric and produces $C^0$  discontinuities between elements on edges.
Ubach et al. \cite{Ubach2010}  present a shell element based on B\'ezier triangles that allows for a better behavior.
However, the B\'ezier interpolation defined on each triangle element is based on the motion computed on adjacent triangles.
So the stiffness forces computed on each element not only depends on nodes of the element but also on all the adjacent nodes. 
This greatly reduces the sparsity of the global stiffness matrix and consequently increases the computation time.

We start by reviewing the definition of B\'ezir triangle. Then the construction
of initial control mesh is described and  kinematic link between control points
is defined. In Section 3 the FE model is derived and the method of computing
the corotational frame is described. In Section 4 we specify how to map a high
resolution surface mesh onto the mechanical mesh and give formulas for
computing velocities of points on the surface and projecting of forces. Section
5 presents test of convergence, speed analysis and two created applications
that use the described shell model. We conclude with discussion of present
limitations.

@@

Extending the work of Comas \cite{Comas2010c} we have designed a shell element that uses B\'ezier interpolation polynomials.
Our element solves the problems of the element presented by Comas:
(i) it uses all 6 degrees of freedom,
(ii) the deformations are completely symmetric and
(iii) the boundary between elements is continuous.
We present a two stage interpolation:
(i) first a kinematic link between 6 DOF nodes of element and nodes of B \'ezier control mesh is defined and
(ii) interpolation function of cubic B\'ezier triangle is applied to interpolate on the surface of the element.
Mapping of high resolution mesh on the curved surface of the shell.
Finally from the kinematic link we derive formulas for
(i) interpolating velocity on the element surface and for
(ii) distribution of force applied on the element surface back to 6 DOF nodes of the element.
This allows us to use smooth mapped surface for collision detection and allows the propagation of proper collision response back onto the mechanical mesh.

@@

In the rest of the document we define the surface over the triangle:
%
\begin{equation}
    (\xi_1,\xi_2) \in \bigtriangleup = \left\{ (\xi_1,\xi_2)~|~\xi_1, \xi_2 \ge 0
        \mathrm{~and~} \xi_1+\xi_2 \le 1 \right\}
\end{equation}
%
B\'ezier triangles were first designed in early sixties by French mathematician de Casteljau.
As with other B\'ezier or B-spline surfaces the B\'ezier triangle is
defined by a mesh of control points that do not necessarily lie on the
surface. The general $n$-th order B\'ezier triangle requires $(n + 1)(n +
2)/2$ control points and it's surface is defined as:
%
\begin{equation}
    T(\xi_1, \xi_2) = \sum_{0 \le i + j \le n} B^n_{i,j}(\xi_1,\xi_2) P_{i,j}
\end{equation}
%
where $P_{i,j}$ are the control points and $B^n_{i,j}$ are the bivariate
Bernstein basis functions defined as:
%
\begin{equation}
  B^n_{i,j} (\xi_1,\xi_2) =
    \binom{n}{i~j} \xi_1^i \xi_2^j (1-\xi_1-\xi_2)^{n-i-j},
\end{equation}
\begin{equation}
  \binom{n}{i~j} = \frac{n!}{i!j!(n-i-j)!}
\end{equation}
%
where $ 0 \le i+j \le n $.

In our work we use cubic B\'ezier triangle ($n=3$). Triangle of smaller order
would fail to describe all possible deformations of the element and
triangles of higher order cannot be described unambiguously with only 18
DOFs available to us without adding further constraints on the control
mesh. Cubic triangle is described by 10 control points and the surface is
explicitly defined as (see fig. \ref{fig-bezier2}):
%
\begin{equation}\label{eq-cubicbez}
  \begin{split}
  T(\xi_1,\xi_2) & =
             \xi_1^3\ P_1
           + \xi_2^3\ P_2
           + \xi_3^3\ P_3 \\
         & + 3\ \xi_1^2 \xi_2\ P_4
           + 3\ \xi_1^2 \xi_3\ P_5
           + 3\ \xi_2^2 \xi_3\ P_6 \\
         & + 3\ \xi_1 \xi_2^2\ P_7
           + 3\ \xi_1 \xi_3^2\ P_8
           + 3\ \xi_2 \xi_3^2\ P_9 \\
         & + 6\ \xi_1 \xi_2 \xi_3\ P_{10} \\
  \end{split}
\end{equation}
%
where $ \xi_3 = 1 - \xi_1 - \xi_2 $. By naming $n_i$ the respective values
of the Bernstein basis functions linked to control point $P_i$ we can
shortly express \eqref{eq-cubicbez} as:
%
\begin{equation}\label{eq-cubicbez2}
    T = \sum_{i=1}^{10} n_i P_i
\end{equation}

@@

The cubic B\'ezier triangles are defined by a mesh of 10 control points each, however the
simulated object is usually described by a triangular mesh possibly with normals.
A procedure for construction of initial B\'zier mesh is necessary.
We construct the initial B\'ezier mesh by taking into account the curvature of the triangular
mesh. Note that special care has to be taken to maintain
the continuity between elements across the nodes and edges, 
we partially employ the method described in \cite{Ubach2010} to maintain
$C^0$ continuity. Each of the control points on the edge is
computed as the intersection of:

@@

Notice that the points on the edge will be same for both triangles sharing
the edge.

$C^1$ continuity between elements on the edges can also be maintained if special care is taken
when computing the position of the central point
(see \cite{Ubach2010} or refer to the work of Farin \cite{Farin2002} for continuity conditions). 
For simplicity we decided to keep the edge boundary only $C^0$ continuous and we compute the central point as a function of other 9 points: 
%
\begin{equation}\label{eq-central}
    P_{10} = \frac{1}{3}(\sum_{i=4}^9 P_i - \sum_{i=1}^3 P_i)
\end{equation}
%
where the points $P_i$ are named as in Figure \ref{fig-bezier2}.

@@

With 10 control points and 3 DOFs per point it is a total of 30 DOFs for triangle.
Simulation of such complex system would significantly reduce the sparsity of stiffness matrix.
We propose a reduction of the number of degrees of freedom by using 3 nodes with 6 degrees of freedom
by creating a kinematic link between the control points.
Each of the edge control points is attached to nearest corner node. See Figure
\ref{fig-segments2} for the correspondence between edge points and
associated corner nodes. 
In the following text the index $j \in [4;10]$ refers to edge points and index $i \in [1;3]$ to the associated corner point/node (unless $i$ is explicitly used differently).

@@

In the initialization phase we remember the position of edge point $j$ in the frame of corner point $i$:
%
\begin{equation}
    \overline{P_{i-j}} = R(\theta_i^0)^T P_{i-j}^0
\end{equation}
%
where
$\mathbf{x}_i$ and $\theta_i$ are the position and angular position of the node $i$,
$R(\theta_i^0)$ is the rotation matrix for the angular position of node $i$ at the beginning of the simulation and
$P_{i-j}^0 = P_j^0 - \mathbf{x}_i^0 = P_j^0 - P_i^0$ the attachment segment.

In each step of the simulation we
%recompute the new position based on current orientation of the corner node:
apply the rigid transformation of frame $i$ on control point $j$:
%
\begin{equation}
    P_j = \mathbf{x}_i + P_{i-j} = P_i + R(\theta_i) \overline{P_{i-j}}
\end{equation}

Similarly for the central point we apply rigid transformation of all frames of the triangle and compute the mean:
%
\begin{equation}\label{eq-central2}
    P_{10} =  \frac{1}{3}\left( \sum_{i=1}^3 \mathbf{x}_i + R(\theta_i) \overline{P_{i-10}} \right)
\end{equation}

Putting previous equations into \eqref{eq-cubicbez2} we get:
%
\begin{equation}\label{global-transform} \begin{split}
    & T = \sum_{i=1}^{10} n_i P_i = \sum_{i=1}^{3} n_i \mathbf{x}_i
    + \sum_{j=4}^9 n_j  \left(
        \mathbf{x}_i + R(\theta_i) \overline{P_{i-j}} \right) + \\
    &\quad + \frac{n_{10}}{3}
        \left( \sum_{i=1}^3 \mathbf{x}_i + R(\theta_i) \overline{P_{i-10}} \right)
    \end{split}
\end{equation}

@@

A small variation $\delta \mathbf{x}_i$ of the node $i$ position and $\delta \mathbf{\theta}_i$ for the rotation produces the following small variation of position corner points $\delta P_i$ and edge points $\delta P_j$:
%
\begin{eqnarray}
    \delta P_i & = & \delta \mathbf{x}_i \\
    \delta P_j & = & \delta \mathbf{x}_i +  P_{i-j} \times \delta \mathbf{\theta}_i
\end{eqnarray}
%
and variation of central point $\delta P_{10}$:
%
\begin{equation}
    \delta P_{10} = \frac{1}{3} \sum_{i=1}^3 \left(
    \delta \mathbf{x}_i + P_{i-10} \times \delta \mathbf{\theta}_i \right)
\end{equation}

Thus, finally, it creates a variation of the point position on the surface $\delta T$:
%
\begin{equation}
    \delta T = \sum_{i=1}^{10} n_i  \delta P_i
\end{equation}
%
or:

\begin{equation}\label{eq-bez-delta}
    \begin{split}
        \delta T = &   (n_1 + n_4 + n_5 + \frac{1}{3} n_{10})  \delta \mathbf{x}_1 + \\
                   & + (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})  \delta \mathbf{x}_2 + \\
                   & + (n_3 + n_8 + n_9 + \frac{1}{3} n_{10})  \delta \mathbf{x}_3 + \\
                   & + \sum_{j=4}^9 n_j P_{i-j} \times \delta \mathbf{\theta}_i + \\
                   & + \frac{n_{10}}{3} \sum_{i=1}^3 P_{i-10} \times \delta \mathbf{\theta}_i 
    \end{split}
\end{equation}

We have simplified the notation by leaving the variable $i$ free in the above sums ($\sum_{j=4}^9$) involving $P_{i-j}$.
Note however that the expressions are sound and unambiguous because there is only one possible $i$ for each $j$ depending on the connection of segments to the corner nodes (see fig. \ref{fig-segments2}).

@@

We define a local system of coordinates $(x,y,z)$ in which the axis z is oriented along the thickness of the element (the computation of the frame of each element will be detailed in section \ref{sec-corot}). 
Given the initial position of the triangle points in this system of coordinate, $\{x_1^0, y_1^0\}$,  $\{x_2^0, y_2^0\}$, and  $\{x_3^0, y_3^0\}$, we can compute a linear relation between $\{\xi_1,\xi_2, \xi_3\}$ and $\{ x,y \}$:
%
\begin{equation}
\label{shape-vector}
\left[ \begin{array}[c]{c} \xi_1(x,y) \\ \xi_2(x,y) \\ \xi_3(x,y) \end{array} \right] =
\left[ \begin{array}[c]{ccc}
1 & 1 &  1 \\
x_1^0 & x_2^0 & x_3^0 \\
y_1^0 & y_2^0 & y_3^0
\end{array} \right] ^{-1} 
\left[ \begin{array}[c]{c} 1 \\ x \\ y \end{array} \right] 
\end{equation}

@@

The term $\deriv{T^k}{\xi_i}$ can be derived from the expression of $T(\xi_1,\xi_2, \xi_3)$ in equation (\ref{global-transform}) and 
$\deriv{\xi_i}{l}$ from equation (\ref{shape-vector}). 
The in-plane displacement gradient can be obtained with tensor $F$:
%
\begin{equation}
\deriv{u^k}{l} = F_{kl} - I_{kl} 
\end{equation}
%
where $\mat{I}$ is the identity matrix ($I_{kl}  = \{1$ if $l=k$ ;  $0$ if $l \neq k \} $). 
The displacement gradient can be linked to the degrees of freedom of the shell using the expression of the displacement of a point $\delta T$ (see equation  (\ref{eq-bez-delta})) on the shell surface:
\begin{equation}
    \nabla u = \nabla(\delta T)
\end{equation}

@@

For the bending formulation, the shell theory uses a measure of the out of plane-displacement $u^z(x,y)$
The strain tensor is based on Kirchhoff-Love theory for thin plates:
\begin{equation}\label{out-plane-strain}
    \begin{bmatrix}
        \epsilon_{xx} \\ \epsilon_{yy} \\ 2\epsilon_{yx}
    \end{bmatrix}
    =
    \begin{bmatrix}
        -z \deriv{^2 u^z}{x^2}  \\[1mm]  -z \deriv{^2 u^z}{y^2}  \\[1mm]  - 2z \deriv{^2 u^z}{xy}
    \end{bmatrix}
\end{equation}
where $z$ is the local coordinate along $z$-axis of the point on which the strain is measured.

>>>>

% --- for  Annual CONFERENCE
% \ConferenceSubmission % uncomment for Conference submission
% \ConferencePaper      % uncomment for (final) Conference Paper
% \STAR                 % uncomment for STAR contribution
% \Tutorial             % uncomment for Tutorial contribution
% \ShortPresentation    % uncomment for (final) Short Conference Presentation
%
% --- for  CGF Journal
% \JournalSubmission    % uncomment for submission to Computer Graphics Forum
% \JournalPaper         % uncomment for final version of Journal Paper
%
% --- for  EG Workshop Proceedings
%\WsSubmission    % uncomment for submission to EG Workshop
\WsPaper         % uncomment for final version of EG Workshop contribution
%
 \electronicVersion % can be used both for the printed and electronic version

@@

%\FigureI{name}{caption}
\newcommand{\FigureI}[2]{%
\begin{figure}[htb]
  \centering
  \input{#1.pdf_tex}
  \caption{\label{fig-#1}#2}
\end{figure}}

@@

\author[T. Golembiovský \& C. Duriez]
       {Tomáš Golembiovský$^{1,2}$ and Christian Duriez$^{2}$\\
        $^1$Faculty of Informatics, Masaryk University, Czech Republic\\
        $^2$INRIA Lille -- Nord Europe and University of Lille, France
       }

@@

\begin{classification} % according to http://www.acm.org/class/1998/
    \CCScat{Computer Graphics}{I.3.5}{Computational Geometry and Object Modeling}{Physically based modeling}.
    %\CCScat{Computer Graphics}{I.3.7}{Three-Dimensional Graphics and Realism}{Animation}.
    %\CCScat{Simulation and Modeling}{I.6.8}{Types of Simulation}{Animation}
    %\CCScat{Life and Medical Sciences}{J.3}{Health}{}
\end{classification}

@@

Thin objects are very frequent in our every-day life. % thin wall objects ?
%Objects with thin structure occur in many forms 
e.g. textiles, paper, leaves etc. 
Such objects are also very common in anatomical or pathological structures for example tubular structures (like blood vessels, colon \ldots), thin bag structures (%vessie, 
Glisson capsule, aneurysms,  \ldots), and also many others (eyes, skin, \ldots). 
In the field of surgical simulation, we try to reproduce or anticipate the mechanical behavior of these structures during the surgery.
Thus, it is very important to capture accurately their physical behavior.
However, in the same time, it is also very important to maintain fast computation rates, in order to be compatible with the targeted applications (interactive simulation for education and for planning).

This problem has been faced by the computer graphics community but for different applications.
For instance, many related works address the problem of the simulation of cloth using for example mass-spring models or bending models \cite{Volino2009,Grinspun2003}. %% Todo Choi2007?
However, these models are discrete so their behavior depends on the mesh and their parameters are not easily measurable.
This is completely acceptable for some computer graphics application, where the parameters can be tuned manually to obtain a realistic animation.
In the case of medical simulation, more physically based modeling is needed.

Methods based on continuum mechanics and namely theory of elasticity gained on popularity in real-time simulations ever since Terzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.
While the methods using finite elements and theory of elasticity are fairly popular in volumetric deformation modeling \cite{Cotin1999,Nesme2006,Miller2007}, they are still frowned upon in the area of thin structures. 
Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
This theory is integrated numerically using Finite Elements \cite{Reddy1993}, but there are two issues for obtaining interactive frame rates: 
(i) the continuum equations require second spatial derivatives that can not be handled with linear interpolation and 
(ii) the final system of equations that describes the mechanics is non-linear if large deformations are present.

The usual argument against non-linear FEM based methods is that they are computationally too expensive. 
This is indeed true if lots of elements are used to discretize the object.
A classical strategy, is to use high-polygonal meshes for rendering or collision detection and response, while maintaining a reasonably low number of elements for the mechanics.
The work of Bouthors et al.\cite{Bouthors2007} is a good application example of this strategy for surface deformations.
But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements with interpolation based on high order polynomials.
This type of interpolation is necessary because bending continuum model
(based on Kirchhoff-Love theory for thin plates)  requires the computation of second spatial derivatives.
On the other hand, elements common in mechanics (e.g. Discrete Kirchhoff Triangles) are based on shape functions that are impractical for real-time simulations or don't have explicitly defined shape functions at all.
In \cite{Comas2010c}, Comas et al. use a hybrid interpolation: linear interpolation is used for in-plane deformations and cubic polynomials for bending deformations.
The element is piloted by 3 nodes on which 6 degrees of freedom are defined. 
From this point of view, our work is derived from this approach but we address several noted drawbacks: the formulation misses one rotational degree of freedom, the cubic interpolation is not symmetric and produces $C^0$  discontinuities between elements on edges.

B\'ezier based shell elements in their general form have been presented
for both triangular \cite{Ainsworth2011} and quadrilateral \cite{Boumechra2006} elements. High number of simulated DOFs makes such formulations impractical for simulations.
Ubach et al. \cite{Ubach2010} presented a rotation-free 3-node shell element based on B\'ezier triangles where the element geometry is estimated from all the neighbouring elements, not just those sharing an edge.
This greatly reduces the sparsity of the global stiffness matrix and consequently increases the computation time.
Also, the lack of rotational degrees of freedom limits the application in surgical simulations.

%Ubach et al. \cite{Ubach2010}  present a shell element based on B\'ezier triangles that allows for a better behavior.
%However, the B\'ezier interpolation defined on each triangle element is based on the motion computed on adjacent triangles.
%So the stiffness forces computed on each element not only depends on nodes of the element but also on all the adjacent nodes. 
%This greatly reduces the sparsity of the global stiffness matrix and consequently increases the computation time.

We start by reviewing the definition of B\'ezir triangle. Then the construction
of initial control mesh is described and  kinematic link between control points
is defined. In Section 3 the FE model is derived and the method of computing
the corotational frame is described. In Section 4 we specify how to map a high
resolution surface mesh onto the mechanical mesh and give formulas for
computing velocities of points on the surface and for projection of forces. Section
5 presents test of convergence, speed analysis and two created applications
that use the described shell model. We conclude with discussion of present
limitations.

@@

Extending the work of Comas \cite{Comas2010c} we have designed a shell element that uses B\'ezier interpolation polynomials.
Our element solves the problems of the element presented by Comas:
(i) it uses all 6 degrees of freedom,
(ii) the deformations are completely symmetric and
(iii) the boundary between elements is continuous.
We present a two stage interpolation:
(i) first a kinematic link between 6 DOF nodes of element and nodes of B \'ezier control mesh is defined and
(ii) interpolation function of cubic B\'ezier triangle is applied to interpolate on the surface of the element.
Finally from the kinematic link we derive equations for
(i) interpolating velocity at arbitrary position on surface and for
(ii) distribution of forces applied on the element surface back to 6 DOF nodes of the element.
This allows us to use smooth mapped surface for collision detection and allows the propagation of proper collision response back onto the mechanical mesh.

@@

In the rest of the document we define the surface over the triangle:
%
\begin{equation}
    (\xi_1,\xi_2) \in \bigtriangleup = \left\{ (\xi_1,\xi_2)~|~\xi_1, \xi_2 \ge 0
        \mathrm{~and~} \xi_1+\xi_2 \le 1 \right\}
\end{equation}
%
%B\'ezier triangles were first designed in early sixties by French mathematician de Casteljau.
As with other B\'ezier or B-spline surfaces the B\'ezier triangle is
defined by a mesh of control points that do not necessarily lie on the
surface. The general $n$-th order B\'ezier triangle requires $(n + 1)(n +
2)/2$ control points and it's surface is defined as:
%
\begin{equation}
    T(\xi_1, \xi_2) = \sum_{0 \le i + j \le n} B^n_{i,j}(\xi_1,\xi_2) P_{i,j}
\end{equation}
%
where $P_{i,j}$ are the control points and $B^n_{i,j}$ are the bivariate
Bernstein basis functions defined as:
%
\begin{equation}
  B^n_{i,j} (\xi_1,\xi_2) =
    \binom{n}{i~j} \xi_1^i \xi_2^j (1-\xi_1-\xi_2)^{n-i-j},
\end{equation}
\begin{equation}
  \binom{n}{i~j} = \frac{n!}{i!j!(n-i-j)!}
\end{equation}
%
where $ 0 \le i+j \le n $.

In our work we use cubic B\'ezier triangle ($n=3$). Triangle of smaller order
would fail to describe all possible deformations of the element and
triangles of higher order cannot be described unambiguously with only 18
DOFs available to us without adding further constraints on the control
mesh. Cubic triangle is described by 10 control points and the surface is
explicitly defined as (see fig. \ref{fig-bezier2}):
%
\begin{equation}\label{eq-cubicbez}
  \begin{split}
  T(\xi_1,\xi_2) & =
      \xi_1^3\ \mathbf{p}_1
      + \xi_2^3\ \mathbf{p}_2
      + \xi_3^3\ \mathbf{p}_3 \\
      & + 3\ \xi_1^2 \xi_2\ \mathbf{p}_4
      + 3\ \xi_1^2 \xi_3\ \mathbf{p}_5
      + 3\ \xi_2^2 \xi_3\ \mathbf{p}_6 \\
      & + 3\ \xi_1 \xi_2^2\ \mathbf{p}_7
      + 3\ \xi_1 \xi_3^2\ \mathbf{p}_8
      + 3\ \xi_2 \xi_3^2\ \mathbf{p}_9 \\
      & + 6\ \xi_1 \xi_2 \xi_3\ \mathbf{p}_{10} \\
  \end{split}
\end{equation}
%
where $ \xi_3 = 1 - \xi_1 - \xi_2 $. By naming $n_i$ the respective values
of the Bernstein basis functions linked to control point $\mathbf{p}_i$ we can
shortly express \eqref{eq-cubicbez} as:
%
\begin{equation}\label{eq-cubicbez2}
    T = \sum_{i=1}^{10} n_i \mathbf{p}_i
\end{equation}

@@

%The cubic B\'ezier triangles are defined by a mesh of 10 control points each, however the
Since the
simulated object is usually described by a triangular mesh,
a procedure for construction of initial B\'zier mesh is necessary.
%We construct the initial B\'ezier mesh by taking into account the curvature of the triangular
%mesh.
Note that special care has to be taken to maintain
the continuity between elements across the nodes and edges, 
we partially employ the method described in \cite{Ubach2010} to maintain
$C^0$ continuity. Each of the control points on the edge is
computed as the intersection of:

@@

%Notice that the points on the edge will be same for both triangles sharing
%the edge.

$C^1$ continuity between elements on the edges can also be maintained if special care is taken
when computing the position of the central point
(see \cite{Ubach2010} or refer to the work of Farin \cite{Farin2002} for continuity conditions). 
For simplicity we keep the edge boundary only $C^0$ and we compute the central point as a function of other 9 points: 
%
\begin{equation}\label{eq-central}
    \mathbf{p}_{10} = \frac{1}{3}(\sum_{i=4}^9 \mathbf{p}_i - \sum_{i=1}^3 \mathbf{p}_i)
\end{equation}
%
%where the points $P_i$ are named as in Figure \ref{fig-bezier2}.

@@

With 10 control points and 3 DOFs per point it is a total of 30 DOFs for element.
We propose a reduction of the number of DOFs by using 3 nodes with 6 degrees of freedom
and a kinematic link between control points.
Each of the edge control points is attached to nearest corner node. See Figure~\ref{fig-segments2} for the correspondence between edge points and
associated corner nodes. 
In the following text the index $j \in [4;10]$ refers to internal points and index $i \in [1;3]$ to the associated corner point/node %(unless $i$ is explicitly used differently).
if not stated otherwise.

@@

In the initialization phase we remember the position of edge point $j$ in the frame of corner point $i$:
%
\begin{equation}
    \overline{\mathbf{p}_{i-j}} = R(\theta_i^0)^T \mathbf{p}_{i-j}^0
\end{equation}
%
where
$\mathbf{x}_i$ and $\theta_i$ are the position and angular position of the node $i$,
$R(\theta_i^0)$ is the rotation matrix for the angular position of node $i$ at the beginning of the simulation and
$\mathbf{p}_{i-j}^0 = \mathbf{p}_j^0 - \mathbf{x}_i^0 = \mathbf{p}_j^0 - \mathbf{p}_i^0$ the attachment segment.
The 0 in superscript denotes value in rest state.

In each step of the simulation we
%recompute the new position based on current orientation of the corner node:
apply the rigid transformation of frame $i$ on control point $j$:
%
\begin{equation}
    \mathbf{p}_j = \mathbf{x}_i + \mathbf{p}_{i-j} = \mathbf{p}_i + R(\theta_i) \overline{\mathbf{p}_{i-j}}
\end{equation}

Similarly for the central point we apply rigid transformation of all frames of the triangle and compute the mean:
%
\begin{equation}\label{eq-central2}
    \mathbf{p}_{10} =  \frac{1}{3} \sum_{i=1}^3 \left( \mathbf{x}_i + R(\theta_i) \overline{\mathbf{p}_{i-10}} \right)
\end{equation}

Substituting previous equations into \eqref{eq-cubicbez2} we get:
%
\begin{equation}\label{global-transform} \begin{split}
    & T = \sum_{i=1}^{10} n_i \mathbf{p}_i = \sum_{i=1}^{3} n_i \mathbf{x}_i
    + \sum_{j=4}^9 n_j  \left(
    \mathbf{x}_i + R(\theta_i) \overline{\mathbf{p}_{i-j}} \right) + \\
    &\quad + \frac{n_{10}}{3}
    \sum_{i=1}^3 \left( \mathbf{x}_i + R(\theta_i) \overline{\mathbf{p}_{i-10}} \right)
    \end{split}
\end{equation}

@@

A small variation $\delta \mathbf{x}_i$ of the node $i$ position and $\delta \mathbf{\theta}_i$ for the rotation produces the following small variation of position corner points $\delta \mathbf{p}_i$ and edge points $\delta \mathbf{p}_j$:
%
\begin{eqnarray}\label{eq-point-rel}
    \delta \mathbf{p}_i & = & \delta \mathbf{x}_i \\
    \delta \mathbf{p}_j & = & \delta \mathbf{x}_i +  \mathbf{p}_{i-j} \times \delta \mathbf{\theta}_i
\end{eqnarray}
%
and variation of central point $\delta \mathbf{p}_{10}$:
%
\begin{equation}\label{eq-point-rel2}
    \delta \mathbf{p}_{10} = \frac{1}{3} \sum_{i=1}^3 \left(
    \delta \mathbf{x}_i + \mathbf{p}_{i-10} \times \delta \mathbf{\theta}_i \right)
\end{equation}

Thus, finally, it creates a variation of the point position on the surface $\delta T$:
%
\begin{equation}\label{eq-bez-delta}
    \delta T = \sum_{i=1}^{10} n_i  \delta \mathbf{p}_i
\end{equation}
%
or:

Substituting \eqref{eq-point-rel}-\eqref{eq-point-rel2} into \eqref{eq-bez-delta} 
one can evaluate the equation in terms of matrix multiplication:

\begin{equation}
    \delta T = \mathbf{J}_t \left[ \delta\mathbf{x}^x_1, \hdots, \delta\mathbf{x}^z_3,
        \delta\theta^x_1, \ldots, \delta\theta^z_3 \right]^T
\end{equation}

%\begin{equation}\label{eq-bez-delta}
%    \begin{split}
%        \delta T = &   (n_1 + n_4 + n_5 + \frac{1}{3} n_{10})  \delta \mathbf{x}_1 + \\
%                   & + (n_2 + n_6 + n_7 + \frac{1}{3} n_{10})  \delta \mathbf{x}_2 + \\
%                   & + (n_3 + n_8 + n_9 + \frac{1}{3} n_{10})  \delta \mathbf{x}_3 + \\
%                   & + \sum_{j=4}^9 n_j P_{i-j} \times \delta \mathbf{\theta}_i + \\
%                   & + \frac{n_{10}}{3} \sum_{i=1}^3 P_{i-10} \times \delta \mathbf{\theta}_i 
%    \end{split}
%\end{equation}
%
%We have simplified the notation by leaving the variable $i$ free in the above sums ($\sum_{j=4}^9$) involving $P_{i-j}$.
%Note however that the expressions are sound and unambiguous because there is only one possible $i$ for each $j$ depending on the connection of segments to the corner nodes (see fig. \ref{fig-segments2}).

@@

We define a local system of coordinates $(x,y,z)$ in which the axis z is oriented along the thickness of the element (the computation of the element frame is covered in section \ref{sec-corot}). 
Given the initial position of the triangle points in this system of coordinate, $\{x_1^0, y_1^0\}$,  $\{x_2^0, y_2^0\}$, and  $\{x_3^0, y_3^0\}$, we can compute a linear relation between $\{\xi_1,\xi_2, \xi_3\}$ and $\{ x,y \}$:
%
\begin{equation}
\label{shape-vector}
\left[ \begin{array}[c]{c} \xi_1(x,y) \\ \xi_2(x,y) \\ \xi_3(x,y) \end{array} \right] =
\left[ \begin{array}[c]{ccc}
1 & 1 &  1 \\
x_1^0 & x_2^0 & x_3^0 \\
y_1^0 & y_2^0 & y_3^0
\end{array} \right] ^{-1} 
\left[ \begin{array}[c]{c} 1 \\ x \\ y \end{array} \right] 
\end{equation}

@@

The term $\deriv{T^k}{\xi_i}$ can be derived from the expression of $T(\xi_1,\xi_2, \xi_3)$ in equation (\ref{global-transform}) and 
$\deriv{\xi_i}{l}$ from equation (\ref{shape-vector}). 
The in-plane displacement gradient can be obtained with tensor $F$:
%
\begin{equation}
\deriv{u^k}{l} = F_{kl} - I_{kl} 
\end{equation}
%
where $\mat{I}$ is the identity matrix ($I_{kl}  = \{1$ if $l=k$ ;  $0$ if $l \neq k \} $). 
The displacement gradient can be linked to the degrees of freedom of the shell using the expression of the displacement of a point $\delta T$ \eqref{eq-bez-delta} on the shell surface:
\begin{equation}
    \nabla u = \nabla(\delta T)
\end{equation}

@@

For the bending formulation, the shell theory uses a measure of the out of plane-displacement $u^z(x,y)$
The strain tensor is based on Kirchhoff-Love theory for thin plates:
\begin{equation}\label{out-plane-strain}
    \begin{bmatrix}
        \epsilon_{xx} \\ \epsilon_{yy} \\ 2\epsilon_{yx}
    \end{bmatrix}
    =
    \begin{bmatrix}
        -z \deriv{^2 u^z}{x^2}  \\[1mm]  -z \deriv{^2 u^z}{y^2}  \\[1mm]  - 2z \deriv{^2 u^z}{xy}
    \end{bmatrix}
\end{equation}
where $z$ is the local coordinate along z-axis of the point for which the strain is measured.
